### 2019-10-01 ch1_introduction/content.tex
update introduction

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011)\cite{horning_rnaworld_2011}. % TODO: expand the description
	}
	\label{fig:rnaworld}
\end{figure}

The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins. Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).

\begin{figure}[htb!]
  \centering\includegraphics[width=.5\linewidth]{fig/central_dogma} % TODO: change image
  \caption{
    \textbf{Central Dogma.} % TODO: expand the description
  }
  \label{fig:centraldogma}
\end{figure}

During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA. The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA). This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons). A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms. This process of alternative splicing is essential to generate over a 100.000 different proteins starting from a rather small number of genes (20.000-25.000).

 The mature mRNA is then transported to the cytoplasm, where it engages with ribosomes to initiate \textbf{translation}. During this highly evolutionary conserved process, a chain of amino acids, known as the protein building blocks, is being synthesised. Each amino acid is specified by three nucleotides (a codon) in the mRNA, according to a nearly universal genetic code. After being released by the ribosomes, the translation product undergoes a variety of chemical modifications to form the final folded protein, the structure of which is determined by the sequence of different amino acids in the chain. In addition, polypeptides may be cleaved to yield more than one active polypeptide product. The structure of a protein determines its functionality, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.
 % Ref: Matera AG, Wang Z. A day in the life of the spliceosome. Nature reviews Molecular cell biology. 2014 Feb;15(2):108. 
 % Ref: Book: Human Molecular Genetics, Strachan and Read

\subsection{Cell types}
\textit{Homo sapiens} like to categorise everything they encounter, and so too have they conceptualised groups of cells called "cell types". The human body contains more than 200 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous. This however, is a major underestimation of the real number of cell types. Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types.
% Ref: Strachan and Read

@@

\subsection{Cell dynamics and gene regulation}
Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality. The variation within and between cells happens because of differences in gene expression. The mechanism by which this happens is called gene regulation, and takes place primarily at the level of transcription. According to the needs of a cell, different genes are being transcribed. Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages.
%ref: Strachan and Read
% CVC

@@

\begin{figure}[htb!]
  \centering
  A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
  % also figure with cell numbers
  C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
  % TODO: change image
  \caption{
    A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}. 
    B. 
    C. scmultiomics\cite{moudgil_multimodalscrnaseq_2019}. 
    % TODO: rewrite the description
  }
  \label{fig:profiling_convergence}
\end{figure}

@@

\section{Computational tools}
The rapidly advancing field of single-cell omics harbours exceptional opportunities to discover new aspects of biology and redefine existing knowledge.

Some of these opportunities lie in efforts such as the Human Cell Atlas. The HCA consortium has set out to redefine all human cell types in terms of their gene expression and location, and the developmental trajectories which connect the different cell types. As part of this endeavour, the consortium will perform single-cell omics on tens or even hundreds of millions of cells.

Single-cell omics permits new types of analyse but also come with hitherto unseen data characteristics, the combination of which poses exciting new challenges for the computational community to tackle (Figure \ref{fig:comp_tools}A)\cite{stegle_computationalanalyticalchallenges_2015,yuan_challengesemergingdirections_2017,chen_singlecellrnaseqtechnologies_2019}. These challenges include:
\begin{itemize}
  \item normalisation: separating biological noise from technical noise,
  \item dimensionality reduction: providing a visual and informative overview of a given dataset, 
  \item trajectory inference: identifying and characterising transitions between different cellular states, and
  \item gene regulatory network inference: inferring regulatory interactions between transcription factors across individual cells.
\end{itemize}

\begin{figure}[htb!]
  \centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
  \centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
  % TODO: change image
  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016)\cite{wagner_revealingvectorscellular_2016}.
    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}
    % TODO: rewrite the description
  }
  \label{fig:comp_tools}
\end{figure}

\mycomment{Make a better connection to the subsections below}

\subsection{Normalisation}
\mycomment{TODO}

>>>>

\begin{figure}[htb!]
	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011)\cite{horning_rnaworld_2011}.
	  \fbox{\color{red} TODO: combine images, rewrite the description.}
	}
	\label{fig:rnaworld}
\end{figure}

The central dogma describes the general flow of genetic information in almost all existing living cells: DNA is decoded to RNA, which in turn encodes proteins\cite{strachan_humanmoleculargenetics_2011}. Main processes involved in the central dogma are \textbf{transcription}, \textbf{splicing}, and \textbf{translation} (Figure \ref{fig:centraldogma}).

\begin{figure}[htb!]
  \centering\includegraphics[width=.5\linewidth]{fig/central_dogma} % TODO: change image
  \caption{
    \textbf{Central Dogma.}
    \fbox{\color{red} TODO: combine images, rewrite the description.}
  }
  \label{fig:centraldogma}
\end{figure}

During the process of \textbf{transcription} that takes place in the cell nucleus, a complementary RNA copy is transcribed from the template DNA. The initial RNA transcript is a precursor messenger RNA (pre-mRNA) that needs to undergo series of maturation steps to ultimately form the mature messenger RNA (mRNA). This maturation includes pre-mRNA \textbf{splicing} to remove non-protein coding intervening sequences (the introns) and to join the neighbouring protein-coding sequences (the exons). A single pre-mRNA can be alternatively spliced to generate multiple forms of mRNAs that will result in the production of multiple protein isoforms. This process of alternative splicing is essential to generate more than 100'000 different proteins starting from just 20'000 genes\cite{nilsen_expansioneukaryoticproteome_2010}.

The mature mRNA is then transported to the cytoplasm, where it engages with ribosomes to initiate \textbf{translation}. During this highly evolutionary conserved process, a chain of amino acids, known as the protein building blocks, is being synthesised. Each amino acid is specified by three nucleotides (a codon) in the mRNA, according to a nearly universal genetic code. After being released by the ribosomes, the translation product undergoes a variety of chemical modifications to form the final folded protein, the structure of which is determined by the sequence of different amino acids in the chain. In addition, polypeptides may be cleaved to yield more than one active polypeptide product. The structure of a protein determines its functionality, which includes catalysing biochemical reactions, providing structure, and transportation of molecules.

\subsection{Cell types}
\textit{Homo sapiens} like to categorise everything they encounter, and so too have they conceptualised groups of cells called "cell types". The human body contains more than 200 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous. This however, is a major underestimation of the real number of cell types. Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types\cite{strachan_humanmoleculargenetics_2011}.

@@

\subsection{Cell dynamics and gene regulation}
Cells are dynamic entities that can gradually produce the molecules needed to acquire new functionality. The variation within and between cells happens because of differences in gene expression. The mechanism by which this happens is called gene regulation, and takes place primarily at the level of transcription. According to the needs of a cell, different genes are being transcribed. Housekeeping genes are being expressed in essentially every cell, while other genes are cell type or tissue specific and may be expressed at certain developmental stages\cite{strachan_humanmoleculargenetics_2011}.

% CVC

@@

\begin{figure}[htb!]
  \centering
  A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
  % also figure with cell numbers
  C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
  \caption{
    A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014)\cite{junker_everycellspecial_2014}. 
    B. Cell numbers vs. gene count over time.
    C. scmultiomics\cite{moudgil_multimodalscrnaseq_2019}. 
    \fbox{\color{red} TODO: combine images, rewrite the description.}
  }
  \label{fig:profiling_convergence}
\end{figure}

@@

\section{Computational tools}
The new types of analyses permitted by single-cell omics harbour exceptional opportunities to discover new aspects of biology and redefine existing knowledge (Figure \ref{fig:comp_tools}A). These include the following categories\cite{zappia_exploringsinglecellrnaseq_2018}.
\begin{itemize}
	\item \textbf{Data imputation}: \ldots.
	\item \textbf{Data integration}: \ldots.
	\item \textbf{Clustering}: \ldots.
	\item \textbf{Dimensionality reduction}: providing a visual and informative overview of a given dataset.
	\item \textbf{Trajectory inference}: identifying and characterising transitions between different cellular states.
	\item \textbf{Trajectory alignment}: \ldots.
	\item \textbf{Network inference}: inferring gene regulatory interactions between transcription factors across individual cells.
\end{itemize}
\mycomment{To do: write a section for each of these categories.}

Developing new computational tools to perform these analyses has proven challenging for two main reasons \cite{stegle_computationalanalyticalchallenges_2015, yuan_challengesemergingdirections_2017, chen_singlecellrnaseqtechnologies_2019}. Firstly, single-cell omics data suffers from hitherto unseen noise characteristics, including high dropout rates and high levels of transcriptional stochasticity. In addition, a low abundance of single-cell omics data that can be used as a ground-truth obstruct quantitative evaluation of such methods.

\begin{figure}[htb!]
  \centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
  \centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
  \caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016)\cite{wagner_revealingvectorscellular_2016}.
    B. Zappia et al. (2018)\cite{zappia_exploringsinglecellrnaseq_2018}.
    \fbox{\color{red} TODO: combine images, rewrite the description.}
  }
  \label{fig:comp_tools}
\end{figure}
